---
title: "Getting Started: Diamonds"
author: "Hannah Sarver"
date: 2020-07-09
output:
  github_document:
    toc: true
---

In this challenge, you will go through the process of exploring, documenting, and sharing an analysis of a dataset. We will use these skills again and again in each challenge.

```{r setup}
library(tidyverse)
```

# Data Exploration
<!-- -------------------------------------------------- -->

In this first stage, you will explore the `diamonds` dataset and document your observations.

__q1__ Create a plot of `price` vs `carat` of the `diamonds` dataset below. Document your observations from the visual.

```{r q1-task}
ggplot(diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

**Observations**:

* There is an apparent trend of covariance that as carat increases, price increases, but with a widening spread of price for carat values at 1 or higher.
* There are large numbers of data points on the whole and half number values of carat, perhaps suggesting that some of the source of the data was biased towards some discrete values.
* There are several outliers in the carat axis but what appears to be a very flat maximum value in price, even among the far outliers in carat.

__q2__ Create a visualization showing variables `carat`, `price`, and `cut` simultaneously. Experiment with which variable you assign to which aesthetic (`x`, `y`, etc.) to find an effective visual.

```{r q2-task}
ggplot(diamonds) +
  geom_smooth(data = subset(diamonds, carat < 2.5), mapping = aes(x = carat, y = price, color = cut)) +
  geom_point(data = subset(diamonds, carat >= 2.5), mapping = aes(x = carat, y = price, color = cut))
```

**Observations**:

* Using smooth gives a clearer sense of the relationship between cuts at lower carat values, but becomes hard to reason about or even misleading above carats of about 2.5 as we have much less data so trends are not so meaningful. Leaving data with carat of 2.5 or higher as a scatter plot shows few points with a lot of variation.
* In general, worse cuts have lower prices for the same carat, with Fair cuts showing a greater difference in price from the other cuts.
* One observation from the higher-carat data left as scatter points in this graph is that the three very highest carat values are all Fair cut. Perhaps the value of such high carat diamonds is deemed too high to waste any by cutting (we could also look at size as a factor here, or just look up specific information about the few diamonds in question).
* There is a leveling off or downturn in the trend of price as carat increases past about 2.0. This appears across all cuts, but may also be affected by wider variation in the smaller total number of samples as carat increases. This also seems consistent with the apparent flat maximum price we observed in the first visualization.

