---
title: "Aluminum Data"
author: "Hannah Sarver"
date: 2020-07-20
output:
  github_document:
    toc: true
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle
<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

__q1__ Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, mu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
    pivot_longer(
      names_to = c(".value", "angle"),
      names_sep = "_",
      cols = c(-thick, -alloy)
    ) %>%
  mutate(angle = as.integer(angle)) %>%
  filter(E > 0)
df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "mu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA
<!-- -------------------------------------------------- -->

## Initial checks
<!-- ------------------------- -->

__q2__ Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own question that you'd like to answer about the data.

```{r q2-task}
##
summary(df_stang_long)
```

**Observations**:

- Is there "one true value" for the material properties of Aluminum?
  - No, properties vary across parameters
- How many aluminum alloys were tested? How do you know?
  - Looks like just one; the summary is not super clear but since it's a small data set we can confirm by glancing through the full set
- What angles were tested?
  - 0, 45, 90
- What thicknesses were tested?
  - 0.022, 0.032, 0.064, 0.081 (units?)
- Which of angle or thickness affects the material properties more? Is the answer different for E versus mu?
- Probably out of scope for data science, but what values would be desirable for E or mu for given engineering work?

## Visualize
<!-- ------------------------- -->

__q3__ Create a visualization to investigate your question from q1 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q1 here
df_stang_long %>%
  ggplot(aes(x = thick, y = E, color = angle)) +
  geom_point() +
  geom_line() +
  facet_grid(. ~ angle)
```

**Observations**:

- The general shape of the curve for E versus thickness looks similar across the three angles measured.
- It appears that the measured value for E jumps around a bit at the lower thickness values and then very steeply drops at the highest thickness of 0.081 in.
- There are not very many data points, so my attempts to look at any smoothed graphs did not seem to show much trend that felt trustworthy, other than the drop at the highest thickness.
- The values of E seem to be on very discrete points, and glancing at the data table shows accuracy only to the nearest 100, so that also makes it a bit harder to see trends along the lower thickness values.
- Overall for E it appears that thickness matters more than angle, but really only shows much difference for the highest thickness value.

```{r q3-taskB}
df_stang_long %>%
  ggplot(aes(x = thick, y = mu, color = angle)) +
  geom_point() +
  geom_line() +
  facet_grid(. ~ angle)

```

**Observations**:

- The curves for mu look somewhat similar to those for E - at a high level the values jump around a bit at lower thicknesses and then drop off more steeply at the highest thickness value, although to a lesser extent than we saw for E.
- An exception to the above observation is one of the points of measured mu at 90 degrees angle and 0.022 in thickness, which appears to be a bit of an outlier (but again is a little hard to reason about since there is not that much data here).
- Overall as with E the angle does not seem to have very much effect on the measured mu, and the thickness does not have much either besides the drop at the highest thickness value.
- Between E and mu, the dip at the highest thickness appears a bit more prominent for E.

__q4__ Consider the following statement:

"A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(mu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

- This graph appears to contradict the statement, since the group of points for the 0.081 in thickness alloy all have distinctly lower measured E values than the other thicknesses, although among the three lower thicknesses there is not a clear relationship.
- I would say this data shows a pretty conclusive contradiction, presuming that the experiment had some measures to control for other property differences in the higher thickness samples (I am not sure I know enough about the production and test to say what, exactly, gotta read some more) that might impact the elasticity measurement.

# References
<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[https://en.wikipedia.org/wiki/List_of_materials_properties]
