---
title: "Michelson Speed-of-light Measurements"
author: "Hannah Sarver"
date: 2020-07-17
output:
  github_document:
    toc: true
---

*Purpose*: When studying physical problems, there is an important distinction
between *error* and *uncertainty*. The primary purpose of this challenge is to
dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure
the speed of light. His approach was a development upon the method of Foucault,
and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second
(in a vacuum). This is very close to the modern *exact* value of `r
LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original
data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk
below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value
`LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in
terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

__q1__ Re-create the following table (from Michelson (1880), pg. 139) using
`df_michelson` and `dplyr`. Note that your values *will not* match those of
Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- 
  df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    MeanVelocity = round(mean(Velocity), -1)
  ) %>%
  ungroup()

df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:

* I rounded the mean to the nearest ten (interpreted in R as round to -1 digit) to match the table from Michelson. I would presume that he rounded his results as an accommodation for both or either of expected error in the data and effects of averaging.
* Looking at the data in the table, it appears that the more distinct measurements agreed more closely with each other than the less distinct, but that the average value of Velocity measured in those observations is further from the exact value than the observations with Distinctness of 1.
  * There are more observations at the higher Distinctness levels, so the distance of the Distinctness 1 average from the others could be from the small sample size rather than something about those measurements actually being more accurate.

The `Velocity` values in the dataset are the speed of light *in air*; Michelson
introduced a couple of adjustments to estimate the speed of light in a vacuum.
In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from
Michelson (1880), pg. 141). While this isn't fully rigorous ($+92$ km/s is based
on the mean temperature), we'll simply apply this correction to all the
observations in the dataset.

__q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to
`Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <-
  df_michelson %>%
  mutate(
    VelocityVacuum = Velocity + 92
  )
df_q2
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

__q3__ Compare Michelson's speed of light estimate against the modern speed of
light value. Is Michelson's estimate of the error (his uncertainty) greater or
less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!
mich_error <- abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON)
mich_error
mich_uncertainty <- LIGHTSPEED_PM
mich_uncertainty
mich_uncertainty_error <- abs(mich_uncertainty - mich_error)
mich_uncertainty_error
```

**Observations**:

* Michelson's uncertainty is much less than his actual error, about 1/3 the value.
* This would imply to me that he either missed or under-accounted for some sources of error.
* To be fair, for values this large and hard to measure, I am still pretty impressed at the accuracy.

__q4__ You have access to a few other variables. Construct a few visualizations
of `VelocityVacuum` against these other factors. Are there other patterns in the
data that might help explain the difference between Michelson's estimate and
`LIGHTSPEED_VACUUM`?

```{r q4-taskA}
## TODO: Visualize VelocityVacuum against variables in the data set
df_q2 %>%
  ggplot() +
  geom_point(
    aes(x = Temp, y = VelocityVacuum, color = Distinctness)
  ) +
  geom_hline(
    aes(yintercept = LIGHTSPEED_VACUUM)
  ) +
  geom_text(
    aes(65, LIGHTSPEED_VACUUM - 20, label = "LIGHTSPEED_VACUUM")
  )
```

**Observations:**

* There does not seem to be any clear relationship between the temperature and measured velocity, the measurements look fairly evenly distributed along both of those axes.
* There also does not seem to be an obvious relationship between the distinctness and velocity, it looks to me like the measurements are more central in the range of measured velocities but this could just be the smaller number of datapoints.
* There are a few further outlying velocity measurements, but these do not appear to be particularly correlated to any temperature range or distinctness level.
* It is interesting that Michelson measured values mostly higher than the real value (even discounting the adjustment for expectation of air vs vacuum).

```{r q4-taskB}
## TODO: Visualize VelocityVacuum against variables in the data set
df_q2 %>%
  ggplot() +
  geom_density(
    aes(VelocityVacuum, group = Distinctness, color = Distinctness)
  )
```

**Observations:**

* Looking at the densities of the velocity vacuum measurements, we can see that they have a mostly normal distribution, with some bumps. What I take from this is that across measurements, the methods gave similar results with normally distributed "noise" in a range with a central peak.
* The highest distinctness shows most "bumpy" shape. I do not have good intuition for why that is.
* The peaks between the three distinctnesses are close together, meaning distinctness does not show a significant difference in where the measurements are centered. This indicates to me that there is really a large offset in the experimental error rather than variation in the data leading to an incorrect result.

```{r q4-taskBAnnotated}
df_q2 %>%
  ggplot() +
  geom_density(
    aes(VelocityVacuum, group = Distinctness, color = Distinctness)
  ) +
  geom_vline(
    aes(xintercept = LIGHTSPEED_VACUUM)
  ) +
  geom_text(
    aes(LIGHTSPEED_VACUUM - 20, 0.004, angle = 90,
        label = "LIGHTSPEED_VACUUM")
  ) +
  geom_vline(
    aes(xintercept = LIGHTSPEED_MICHELSON)
  ) +
  geom_text(
    aes(LIGHTSPEED_MICHELSON - 20, 0.002, angle = 90,
        label = "LIGHTSPEED_MICHELSON")
  ) +
  geom_vline(
    aes(xintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM)
  ) +
  geom_vline(
    aes(xintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM)
  )
```

**Observations:**

* I added some annotations on the same graph as above to show where the Michelson and real value fall, with the expected error range on the Michelson value.
* We do see that Michelson's calculated mean pulls a bit higher than the peaks of the distributions of his (adjusted) measurements, which appears to me to be due to the 2 and 3 distinctnesses having bumps towards the higher side which will pull the average up.
* The measurements still obviously center much higher than the real value.


```{r q4-taskC}
## TODO: Visualize variables in the data set
df_q2 %>%
  ggplot() +
  geom_point(
    aes(x = Date, y = Temp, color = Distinctness)
  )
```

**Observations:**

* I wanted to do a quick check of my expectations that both temperatures and distinctnesses would be clustered based on the date of each set of experiments. There are some clusters but also some single points, I would guess experiments run on their own rather than rapid changes in temperature.
* My takeaway here is that it does not appear that date would be very interesting to look at in terms of any impact independent of the other factors.

## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
