---
title: "Antibiotics"
author: "Hannah Sarver"
date: 2020-07-28
output:
  github_document:
    toc: true
---

*Purpose*: To create an effective visualization, we need to keep our *purpose* firmly in mind. There are many different ways to visualize data, and the only way we can judge efficacy is with respect to our purpose.

In this challenge you'll visualize the same data in two different ways, aimed at two different purposes.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!


```{r setup}
library(tidyverse)
library(ggrepel)

```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization
<!-- -------------------------------------------------- -->

## Purpose: Compare Effectiveness
<!-- ------------------------- -->

__q1__ Create a visualization of `df_antibiotics` that helps you to compare the effectiveness of the three antibiotics across all the bacteria reported. Can you make any broad statements about antibiotic effectiveness?

```{r q1-task}
## TASK: Create your visualization
df_bacteria_grp <- df_antibiotics %>%
  pivot_longer(cols = c(-bacteria, -gram), names_to = "antibiotic")

df_bacteria_grp %>%
  ggplot() +
  #geom_boxplot(aes(antibiotic, value)) +
  geom_point(aes(antibiotic, value, color = gram)) +
  scale_y_log10() +
  geom_hline(yintercept = 0.1) +
  geom_text(aes(0.7, 0.2, label = "MIC = 0.1"))
```

**Observations**:

- Most antibiotic-bacteria pairs appear to have MIC greater than acceptable for human treatment.
- All three antibiotics show a range of MIC values across the bacteria, but penicillin has some very far outliers with super high values (prompting the log scale y axis).
- More of the bacteria under MIC 0.1 for any antibiotic have positive gram than negative, and all of those for penicillin and streptomycin. But there is not an obvious pattern that positive gram show lower MIC than negative generally, although for penicillin it seems that way.
- Out of these three antibiotics, neomycin has the most bacteria with MIC at or under 0.1, although not by a whole lot. I am not sure it is really enough to say that that antibiotic is much more effective than the others in general.

## Purpose: Categorize Bacteria
<!-- ------------------------- -->

The *genus* of a living organism is a human categorization, based on various characteristics of the organism. Since these categories are based on numerous factors, we will tend to see clusters if we visualize data according to relevant variables. We can use these visuals to categorize observations, and to question whether given categories are reasonable!

__q2__ Create a visualization of `df_antibiotics` that helps you to categorize bacteria according to the variables in the data. Document your observations on how how clusters of bacteria in the variables do---or don't---align with their *genus* classification.

```{r q2-task}
## TASK: Create your visualization
df_neomycin <- df_antibiotics %>%
  mutate(bacteria = fct_reorder(bacteria, neomycin))

df_penicillin <- df_antibiotics %>%
  mutate(bacteria = fct_reorder(bacteria, penicillin))

df_streptomycin <- df_antibiotics %>%
  mutate(bacteria = fct_reorder(bacteria, streptomycin))

p1 <- df_antibiotics %>%
  ggplot() +
  geom_point(df_neomycin, mapping = aes(bacteria, neomycin, color = gram)) +
  scale_y_log10() +
  geom_hline(yintercept = 0.1) +
  geom_text(aes(1.6, 0.2, label = "MIC = 0.1")) +
  theme(axis.text.x = element_text(angle = 90))

p2 <- df_antibiotics %>%
  ggplot() +
  geom_point(df_penicillin, mapping = aes(bacteria, penicillin, color = gram)) +
  scale_y_log10() +
  geom_hline(yintercept = 0.1) +
  geom_text(aes(1.6, 0.2, label = "MIC = 0.1")) +
  theme(axis.text.x = element_text(angle = 90))

p3 <- df_antibiotics %>%
  ggplot() +
  geom_point(df_streptomycin, mapping = aes(bacteria, streptomycin, color = gram)) +
  scale_y_log10() +
  geom_hline(yintercept = 0.1) +
  geom_text(aes(1.6, 0.2, label = "MIC = 0.1")) +
  theme(axis.text.x = element_text(angle = 90))

#grid.arrange(p1, p2, p3)
p1
p2
p3
```

**Observations**:

- I chose to look for patterns sorting bacteria by MIC within each antibiotic, and seeing if anything jumped out by genus.
- For neomycin:
  - It appears to be most effective for the two staphylococcus bacteria, also effective for the two salmonella and some others.
  - The streptococcus bacteria are not in agreement on its effectiveness, with two having the highest MIC values and one right on the 0.1 line.
- For penicillin:
  - It is effective for some of the same bacteria as neomycin is. Both staphylococcus, and bacillus.
  - It appears more effective for streptococcus than neomycin is, with one over the 0.1 line but not so far over.
  - Salmonella show similar MIC to each other (on log scale y axes).
  - We can again see clearly the split by gram, and the very far outlying MIC values where penicillin appears very ineffective for some bacteria.
  - Interesting gap between low and high MIC values, with no bacteria on or above MIC = 0.1 until 1.0.
- For streptomycin:
  - Only shows MIC at or below 0.1 for a few bacteria, as we observed above. These include some of the same bacteria as the other antibiotics (staphylococcus and bacillus).
  - I might have guessed from the name that it would be effective against streptococcus, but those show some of the highest MIC values.
  - Shows a relatively even spread (linear-ish on a log scale) relative to the other two where we see some plateaus of groups of bacteria with close MIC values.
- Overall, there are a few patterns:
  - The two staphylococcus bacteria and bacillus are treatable by any of the three antibiotics shown.
  - Some bacteria are not treatable by any of the antibiotics.
  - Not all genus groups show consistent behavior in response to any given antibiotic.

# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." /American Scientist/ (2009) [link](https://www.americanscientist.org/article/thats-funny)
