---
title: "RMS Titanic"
author: "Hannah Sarver"
date: 2020-07-14
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

5 columns:
* Class
* Sex
* Age
* Survived
* N (number in category)

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

* Wikipedia lists an estimated 2224 passengers and crew, versus the total in this dataset is 2201, leaving 23 missing. Possibly it is not known whether or not they survived, or some other column was unknown so they were unlisted.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
    ggplot() +
    geom_col(aes(x = Class, y = n, fill = Sex))
```

**Observations**:

* As I am sure we are about to explore in an upcoming section, this visualizes the total counts but it is difficult to draw any comparisons between the Classes or Sexes because these are raw counts rather than accounting for each group's proportions of the total.
* It does prompt us to look towards the proportional data, because if we naively assumed that there were equal numbers in each Class and Sex, there would be some strange patterns shown. For instance more people survived from 3rd class than 2nd - if there were the same number of passengers in each class that might be a surprising outcome.
* Even without knowing the proportion of the total, we still notice some trends in the data breakdown by Sex. In the higher classes, more female passengers survived (could be that there were more female passengers, or they had a higher rate of survival), in 3rd it appears roughly equal, and in Crew far more males survived than females (again, seeing the proportional breakdown will clarify but my guess would be there were simply far more male crew members in total).

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-taskA}
df_prop %>%
  filter(Survived == "Yes") %>%
    ggplot() +
    geom_col(aes(x = Class, y = Prop, fill = Sex)) +
    ggtitle("Inaccurate Representation of Survival Proportions")
```

**Observations [Unmodified Substitution Replication]**:

If I directly replicate my earlier visualization replacing n with Prop we get a graph that does not seem to make much sense:

* Firstly a stacked plot that showed a total count provides an intuitive comparison between the Classes, but stacking values for a proportion does not seem like an accurate way to make any comparison.
* Also, because the Prop values were calculated accounting for the two Age groups separately, we will end up adding two Prop values per Class/Sex combination, so the resulting visual can show a "proportion" of survivors greater than 1 even for the Class/Sex combo, which is meaningless.

These limitations prompted me to generate the following graph that more accurately depicts the survivor proportion of each Class/Sex group.

```{r q4-taskB}
df_prop_all_ages <-
  df_titanic %>%
  group_by(Class, Sex) %>%
  mutate(
    Total = sum(n),
    Prop_all_ages = n / Total
  ) %>%
  ungroup()
df_prop

df_prop_all_ages %>%
  filter(Survived == "Yes") %>%
    ggplot() +
    geom_col(aes(x = Class, y = Prop_all_ages, fill = Sex),
             position = "dodge") +
    ggtitle("Accurate Representation of Survival Proportions")
```

**Observations**:

* When we group both Adults and Children from a given Class/Sex grouping and separate the groupings into columns, our data makes more sense from a "proportion" perspective. Each column shows the proportion of that population that survived, not a stack of separate proportions.

Now we can make some observations about these data.

* In every group, women survived at a much higher rate than men. A benefit of sexism, I guess?
* The women's survival rate steeply declines with Class, but interestingly the men's survival rate is lowest in 2nd Class rather than in 3rd.
* The men's survival rate in the Crew was a bit higher than the 2nd and 3rd class passengers and nearly as high as even the 1st class men - perhaps once they'd helped the women off they prioritized their own safety next.

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
    geom_col(aes(x = Age, y = Prop, fill = Sex),
             position = "dodge") +
    facet_grid(. ~ Class)
```

**Observations**:

* We now clearly see some important Age-based differences:
  * All the children survived from 1st and 2nd Class
  * There are no children in the crew - not too surprising
  * The children's survival rate in 3rd Class is significantly lower than in 1st and 2nd (this is no doubt somewhat impacted by the very small number of children in the top two classes, as well as the preference given to the richer passengers)
* The "women and children first" philosophy is much more apparent in the difference in survival rate between Adult and Child males.
* We still are missing some information in this view because we only see the proportion of survival within the group, not the relative proportion within each group out of the total passenger/crew count. This means that groups with small totals (ie Female Adult Crew, Children in 1st/2nd Class) will have survival rates more impacted by few individuals.
* This visualization helps us compare survival rates between groups, but does not give us much sense of the numbers of survivors and deceased.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
